---
title: "Comparing weighted vs unweighted"
format: html
editor: visual
---

```{r message = FALSE, warning = FALSE}
.libPaths(c("/enna/users/stfour/shreyar/Desktop/lib/", "/enna/nobackup/biostat/Rpackages/v4/"))
suppressPackageStartupMessages({
  library(lme4)
  library(lmerTest)
  library(spicyR)
  library(SpatialDatasets)
  library(tidySingleCellExperiment)
  library(tidyverse)
  library(scClassify)
  library(caret) 
  library(spatstat.geom)
  library(data.table)
  library(Statial)
  library(BiocParallel)
})

theme_set(theme_classic())
```

```{r}
load(".RData")
```

## Checking scClassify performance

```{r}
# overall accuracy
mean(speSub$cellType == speSub$predLabels)
```

```{r}
# breakdown by cell type... yikes
accuracyDf = speSub |> colData() |> as.data.frame() |>
  dplyr::group_by(cellType) |>
  dplyr::summarise(n = dplyr::n(),
    accuracy = mean(cellType == predLabels)) |>
  dplyr::arrange(accuracy)

accuracyDf
```

```{r}
speSub |> colData() |> as.data.frame() |>
  filter(predLabels == "CD4+ T cells") |> pull(confidenceScore) |> hist()
```


## Comparing weighted vs unweighted coefficients

```{r}
comparison = dplyr::full_join(noWeightsModels |> dplyr::rename(estimate_unweighted = estimate, p_unweighted = p.value,
                                                               p.adj_unweighted = p.adj),
                              weightsModels |> dplyr::rename(estimate_weighted = estimate, p_weighted = p.value,
                                                             p.adj_weighted = p.adj),
                              by = c("from", "to")) |>
  dplyr::select(from, to, estimate_unweighted, estimate_weighted, p_unweighted, p.adj_unweighted, p_weighted, p.adj_weighted) |>
  dplyr::arrange(from, to)    

p = comparison |> 
  dplyr::filter(!is.na(estimate_unweighted), !is.na(estimate_weighted)) |>
  dplyr::mutate(flipped = sign(estimate_unweighted) != sign(estimate_weighted),
                label = paste0(from, " → ", to, "<br>Unweighted: ", round(estimate_unweighted, 3), "<br>Weighted: ",   round(estimate_weighted, 3))) |>
  ggplot(aes(x = estimate_unweighted, y = estimate_weighted, colour = flipped, text = label)) + 
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", alpha = 0.5) +
  scale_colour_manual(values = c("FALSE" = "grey60", "TRUE" = "red")) + 
  labs(x = "Coefficient (unweighted)", y = "Coefficient (weighted)", colour = "Direction flipped") +
  theme_classic()

plotly::ggplotly(p, tooltip = "text")
```

Cell type pairs whose coefficients got flipped:

```{r}
comparison |> filter(estimate_weighted * estimate_unweighted < 0) |> 
  arrange(desc(abs(estimate_weighted - estimate_unweighted)))
```

## Accuracy vs L-function

```{r fig.height = 4, fig.width = 12}
p1 = accuracyDf |>
    dplyr::rename(child1 = cellType, accuracy_child1 = accuracy) |>
    dplyr::inner_join(combinedLfunc, by = "child1") |>
    ggplot(aes(x = accuracy_child1, y = L)) +
    geom_point(alpha = 0.3) +
    geom_smooth(method = "lm") +
    labs(x = "Accuracy (from cell type)", y = "L-function", title = "Unweighted L-function") +
    theme_classic()

p2 = accuracyDf |>
    dplyr::rename(child1 = cellType, accuracy_child1 = accuracy) |>
    dplyr::inner_join(combinedLfunc, by = "child1") |>
    ggplot(aes(x = accuracy_child1, y = weightedL)) +
    geom_point(alpha = 0.3) +
    geom_smooth(method = "lm") +
    labs(x = "Accuracy (from cell type)", y = "L-function", title = "Weighted L-function") +
    theme_classic()

gridExtra::grid.arrange(p1, p2, nrow = 1)
```


## Comparing weighted vs unweighted L-function

```{r}
combinedLfunc = Lfunc
combinedLfunc$weightedL = LfuncW$L

p = ggplot(combinedLfunc |>
             dplyr::mutate(label = paste0(child1, " → ", child2, "\n",
                                                         "L: ", L, "\n", "weighted L: ", weightedL, "\n",
                                                         "imageID: ", imageID)), 
       aes(x = L, y = weightedL, text = label, colour = child1 %in% lowAccuracy), size = 0.5) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", alpha = 0.5) +
  labs(colour = "low accuracy cell type") + 
  theme_classic()

plotly::ggplotly(p, tooltip = "text")
```


### SPP1+ Macrophages

```{r}
p = ggplot(combinedLfunc |> filter(child1 == "SPP1+ Macrophages") |>
             dplyr::mutate(label = paste0(child1, " → ", child2, "\n",
                                                         "L: ", L, "\n", "weighted L: ", weightedL, "\n",
                                                         "imageID: ", imageID)), 
       aes(x = L, y = weightedL, text = label), size = 0.5) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", alpha = 0.5) +
  labs(colour = "low accuracy cell type") + 
  theme_classic()

plotly::ggplotly(p, tooltip = "text")
```


```{r}
tmp = speSub |> colData() |> as.data.frame() |> filter(predLabels == "SPP1+ Macrophages")
table(tmp$predLabels, tmp$imageID)
```

```{r}
tmp = speSub |> colData() |> as.data.frame() |> filter(cellType == "SPP1+ Macrophages")
table(tmp$cellType, tmp$imageID)
hist(tmp$confidenceScore, main = "Distribution of confidence scores for SPP1+ Macrophages", xlab = "confidence score")
```

```{r}
speSub |> colData() |> as.data.frame() |> filter(cellType == "SPP1+ Macrophages") |> pull(predLabels) |> table()
```


These cells got confidently misclassified...

### Mast → Plasma 

```{r}
comparison |> filter(from == "Mast" & to == "Plasma")
```

```{r}
Lfunc$resolution_broad |> levels()
```

```{r}
LfuncW$resolution_broad |> levels()
```

```{r}
ASQW1 
```


```{r}
p = ggplot(combinedLfunc |> filter(child1 == "Mast" & child2 == "Plasma") |>
             dplyr::mutate(label = paste0(child1, " → ", child2, "\n",
                                                         "L: ", L, "\n", "weighted L: ", weightedL, "\n",
                                                         "imageID: ", imageID)), 
       aes(x = L, y = weightedL, text = label), size = 0.5) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", alpha = 0.5) +
  labs(colour = "low accuracy cell type") + 
  theme_classic()

plotly::ggplotly(p, tooltip = "text")
```

Two images of interest: Patient_5_4 and Patient_15_1.

For Patient_5_4, the weighting shows avoidance vs unweighted shows co-localisation. 

```{r}
tmp = speSub |> colData() |> as.data.frame()

 ggplot() +
   geom_point(data = tmp |> filter(imageID == "Patient_5_4" & !predLabels %in% c("Mast", "Plasma")),
               aes(x = x, y = y), colour = "grey", alpha = 0.3) +
   geom_point(data = tmp |> filter(imageID == "Patient_5_4" & predLabels %in% c("Mast", "Plasma")),
              aes(x = x, y = y, colour = predLabels, alpha = confidenceScore)) +
   scale_colour_manual(values = c("Mast" = "red", "Plasma" = "blue")) +
   scale_alpha_continuous(range = c(0.1, 1)) +
   scale_y_reverse() +
   coord_equal() +
   labs(x = "", y = "", title = "Patient_5_4", colour = "Cell type", alpha = "Confidence score")
```

For Patient_15_1, unweighted L-function shows no real interaction vs weighted L shows avoidance. 

```{r}
 ggplot() +
   geom_point(data = tmp |> filter(imageID == "Patient_15_1" & !predLabels %in% c("Mast", "Plasma")),
               aes(x = x, y = y), colour = "grey", alpha = 0.3) +
   geom_point(data = tmp |> filter(imageID == "Patient_15_1" & predLabels %in% c("Mast", "Plasma")),
              aes(x = x, y = y, colour = predLabels, alpha = confidenceScore)) +
   scale_colour_manual(values = c("Mast" = "red", "Plasma" = "blue")) +
   scale_alpha_continuous(range = c(0.1, 1)) +
   scale_y_reverse() +
   coord_equal() +
   labs(x = "", y = "", title = "Patient_15_1", colour = "Cell type", alpha = "Confidence score")
```

### CD4+ T cells → pDC

```{r}
p1 = p1 = combinedLfunc |> filter(child1 == "CD4+ T cells" & child2 == "pDC") |>
  ggplot(aes(x = resolution_broad, y = L)) +
  geom_boxplot() +
  geom_jitter() +
  labs(title = "Unweighted L-function", y = "L", x = "Resolution")

p2 = combinedLfunc |> filter(child1 == "CD4+ T cells" & child2 == "pDC") |>
  ggplot(aes(x = resolution_broad, y = weightedL)) +
  geom_boxplot() +
  geom_jitter() +
  labs(title = "Weighted L-function", y = "weightedL", x = "Resolution")

gridExtra::grid.arrange(p1, p2, nrow = 1)
```



bootstrapping, classifier for T cell subsets 
repeatedly bootstrap each cell and the molecules within each cell with replacement, chuck em into classifier, and get the label. keep doing that, use that instability as a confidence score