---
title: "IMC"
format: html
editor: visual
---

## Libraries and setup

```{r}
.libPaths("/enna/nobackup/biostat/Rpackages/v4/")

suppressPackageStartupMessages({
  library(tidyverse)
  library(tidySingleCellExperiment)
  library(cytomapper)
  library(EBImage)
  library(BiocParallel)
})
```

```{r}
library(imcdatasets, lib.loc = "/dskh/nobackup/shreyar/devPackages/")
damondSPE = Damond_2019_Pancreas(data_type = "spe",
                                 version = "latest",
                                 metadata = FALSE,
                                 on_disk = FALSE,
                                 h5FilesPath = NULL,
                                 force = FALSE)
```

```{r}
damondSPE = readRDS("./data/IMC/damondSPE.rds")
```


```{r}
images = Damond_2019_Pancreas(data_type = "images",
                                 version = "latest",
                                 metadata = FALSE,
                                 on_disk = FALSE,
                                 h5FilesPath = NULL,
                                 force = FALSE)
```

**Islet / Endocrine Cell Identity Markers**

-   INS (Insulin): Produced by beta cells; the primary glucose-lowering hormone and the central autoantigen in T1D. Loss of INS+ cells is the defining feature of disease progression.
-   GCG (Glucagon): Produced by alpha cells; raises blood glucose. Alpha cells are generally preserved in T1D even as beta cells are destroyed.
-   SST (Somatostatin): Produced by delta cells; paracrine regulator of both insulin and glucagon secretion.
-   PPY (Pancreatic Polypeptide): Produced by gamma (PP) cells; regulates pancreatic exocrine secretion and appetite.
-   PDX1 (Pancreatic and Duodenal Homeobox 1): Master transcription factor for beta cell identity and function; required for insulin gene expression and beta cell survival.
-   NKX6_1 (NK6 Homeobox 1): Beta cell-specific transcription factor; co-expressed with PDX1 and used to confirm mature beta cell identity.
-   PCSK2 (Proprotein Convertase 2): Enzyme involved in processing proglucagon and other prohormones; marks alpha and delta cells.
-   IAPP (Islet Amyloid Polypeptide / Amylin): Co-secreted with insulin by beta cells; aggregates to form amyloid deposits in type 2 diabetes.
-   PTPRN (Protein Tyrosine Phosphatase Receptor N / IA-2): Beta cell autoantigen; a major target of autoantibodies in T1D and a marker of neuroendocrine secretory granules.
-   PIN (Proinsulin): Precursor to insulin; marks beta cells with active insulin biosynthesis, and elevated proinsulin:insulin ratio indicates beta cell stress.
-   SYP (Synaptophysin): Pan-neuroendocrine/secretory granule marker; labels all islet cell types and neuroendocrine cells broadly.

**Exocrine Pancreas Markers**

-   AMY2A (Amylase 2A): Produced by acinar cells; digestive enzyme marking the exocrine compartment.
-   KRT19 (Keratin 19): Marks ductal epithelial cells; also expressed in centroacinar cells and can mark ductal metaplasia.
-   CA9 (Carbonic Anhydrase 9): Hypoxia-inducible marker; expressed in ductal cells under stress and strongly upregulated in hypoxic tumors (less relevant here but marks stressed ducts).

**Immune Cell Identity Markers**

-   CD3e: Pan-T cell marker; identifies all T lymphocytes infiltrating the islet (insulitis).
-   CD4: Helper T cell marker; CD4+ T cells coordinate the autoimmune response in T1D.
-   CD8a: Cytotoxic T cell marker; CD8+ T cells are the primary effectors directly killing beta cells in T1D.
-   FOXP3: Transcription factor defining regulatory T cells (Tregs); Treg dysfunction is implicated in failure to suppress autoimmunity in T1D.
-   CD45 (PTPRC): Pan-leukocyte marker; identifies all immune cells regardless of lineage.
-   CD45RA: Marks naive and terminally differentiated effector T cells; helps distinguish naive from memory T cell populations.
-   CD20 (MS4A1): Pan-B cell marker; B cells contribute to T1D through autoantibody production and antigen presentation.
-   CD68: Pan-macrophage marker; macrophages are among the earliest infiltrating cells in insulitis.
-   MPO (Myeloperoxidase): Marks neutrophils and their precursors; neutrophils are increasingly recognized as early contributors to islet inflammation.
-   CD38: Expressed on plasma cells, activated T cells, and NK cells; in this context likely marking activated immune cells within insulitis.
-   PD_1 (CD279): Inhibitory receptor on exhausted/activated T cells; relevant to understanding whether infiltrating T cells are being regulated or are actively cytotoxic.

**Vascular & Stromal Markers**

-   CD31 (PECAM-1): Endothelial cell marker; labels blood vessels and allows assessment of islet vascularization.
-   CDH1 (E-Cadherin): Epithelial cell adhesion molecule; marks ductal and acinar epithelium and helps define tissue architecture.
-   CD44L: Adhesion and signaling molecule expressed on activated/memory immune cells and some stromal cells; marks immune cell activation states.
-   SLC2A1 (GLUT1): Glucose transporter 1; expressed on endothelial cells and upregulated under metabolic stress and hypoxia.
-   CD99: Cell surface marker expressed broadly on leukocytes and some epithelial cells; involved in T cell adhesion and transendothelial migration.

**Proliferation & Cell Cycle Markers**

-   Ki67: Nuclear marker of active proliferation (all phases except G0); used to assess beta cell regeneration or immune cell expansion.
-   p_HH3 (Phospho-Histone H3): Marks cells specifically in mitosis (M phase); more precise proliferation marker than Ki67.
-   p_Rb (Phospho-Retinoblastoma protein): Marks G1/S transition; proxy for cells entering the cell cycle.
-   cPARP_cCASP3 (Cleaved PARP / Cleaved Caspase-3): Dual marker of apoptosis; critical for quantifying beta cell death during autoimmune attack.

**Nuclear / Segmentation Markers**

-   H3 (Histone H3): Nuclear marker for all nucleated cells; used alongside DNA channels for cell segmentation.
-   DNA1 / DNA2: Iridium intercalators (191Ir / 193Ir); stain all nuclei and are the primary markers used for cell segmentation.

```{r}
masks = Damond_2019_Pancreas(data_type = "masks",
                                 version = "latest",
                                 metadata = FALSE,
                                 on_disk = FALSE,
                                 h5FilesPath = NULL,
                                 force = FALSE)
```

```{r eval = F}
damondSPE$x = spatialCoords(damondSPE)[, 1]
damondSPE$y = spatialCoords(damondSPE)[, 2]
```


## Calculate marker variance for each cell

```{r eval = F}
spe = measureObjects(mask = masks,
                           image = images, 
                           img_id = "image_name",
                           return_as = "spe",
                           feature_types = "basic",
                           basic_feature = "sd",
                           BPPARAM = BiocParallel::MulticoreParam(workers = 20))
```

```{r}
# exprs looks log-transfored
hist(assay(damondSPE, "exprs")[4, ])
```

```{r}
# normalised
hist(assay(damondSPE, "quant_norm")[4, ])
```

```{r eval = F}
varMat = assay(spe, "counts_sd")^2
assay(damondSPE, "var", withDimnames = FALSE) = varMat
saveRDS(damondSPE, "./data/IMC/damondSPE.rds")
```

## Calculate pairwise marker correlations


```{r}
computePixelCorrelations = function(images, masks, 
                                    method = "pearson",
                                    minPixels = 3,
                                    BPPARAM = BiocParallel::SerialParam()) {
  
  chNames = channelNames(images)
  nChannels = length(chNames)
  
  # precompute pair names once
  dummyMat = matrix(0, nChannels, nChannels, dimnames = list(chNames, chNames))
  idx = which(upper.tri(dummyMat), arr.ind = TRUE)
  pairNames = paste(chNames[idx[, 1]], chNames[idx[, 2]], sep = "_vs_")
  
  # process one image — will be parallelised across images
  processImage = function(imgName) {
    
    imgArr  = imageData(images[[imgName]])   # [x, y, channels]
    maskArr = imageData(masks[[imgName]])    # [x, y]
    cellIDs = setdiff(unique(as.vector(maskArr)), 0)
    
    # process one cell
    processCell = function(cell) {
      cellPixels = which(maskArr == cell, arr.ind = TRUE)
      if (nrow(cellPixels) < minPixels) return(NULL)
      
      pixelMat = sapply(seq_len(nChannels), function(ch) {
        imgArr[cbind(cellPixels, ch)]
      })
      colnames(pixelMat) = chNames
      
      corMat = suppressWarnings(cor(pixelMat, method = method))
      setNames(corMat[upper.tri(corMat)], pairNames)
    }
    
    cellResults = lapply(cellIDs, processCell)
    names(cellResults) = paste(imgName, cellIDs, sep = "_")
    
    # drop NULLs (cells below minPixels threshold)
    cellResults[!sapply(cellResults, is.null)]
  }
  
  # parallelise across images
  imageResults = bplapply(names(images), processImage, BPPARAM = BPPARAM)
  
  # flatten nested list and bind into matrix
  allResults = do.call(c, imageResults)
  do.call(rbind, allResults)
}
```

```{r eval = F}
corMat = computePixelCorrelations(images = images,
  masks = masks,
  method = "pearson",
  minPixels = 3,
  BPPARAM = BiocParallel::MulticoreParam(workers = 20))

saveRDS(corMat, "./data/IMC/corMat.rds")
```

```{r}
corMat = readRDS("./data/IMC/corMat.rds")
```


```{r}
dim(corMat)
head(corMat[, 1:5])
```

## CD3e vs CD20

### Variance

```{r}
rownames(damondSPE)
```


```{r}
tmp1 = damondSPE |> join_features(features = c("CD3e", "CD20"), shape = "wide", assay = "var") |> 
  colData() |> as.data.frame() |> dplyr::rename(c(var_CD3e = CD3e, var_CD20 = CD20))

tmp2 = damondSPE |> join_features(features = c("CD3e", "CD20"), shape = "wide", assay = "counts") |> 
  colData() |> as.data.frame() |> dplyr::rename(c(mean_CD3e = CD3e, mean_CD20 = CD20))

tmp = cbind(tmp1, tmp2[, c("mean_CD20", "mean_CD3e")])

capVal = function(x, p = 0.99) pmin(x, quantile(x, p, na.rm = TRUE))

tmp = tmp |> mutate(mean_CD3e_cap = capVal(mean_CD3e),
    mean_CD20_cap = capVal(mean_CD20),
    var_CD3e_cap  = capVal(var_CD3e),
    var_CD20_cap  = capVal(var_CD20))

tmp = tmp |> mutate(cell_label = case_when(cell_type == "B" ~ "B",
                                     cell_type %in% c("Tc", "Th") ~ "T",
                                     TRUE ~ "other"))

ggplot(tmp, aes(x = mean_CD3e, y = mean_CD20, colour = var_CD3e_cap)) +
  geom_point(alpha = 0.8, size = 1.2) +
  scale_colour_distiller(palette = "RdYlBu", name = "Var(CD3e)") +
  theme_classic() +
  labs(x = "Mean CD3e", y = "Mean CD20", title = "All cells coloured by variance of CD3e (capped)")

tmp |>
  # filter(cell_label == "other") |>
  ggplot(aes(x = mean_CD3e, y = mean_CD20, 
             colour = capVal(cv_CD3e))) +
  geom_point(alpha = 1, size = 1.2) +
  scale_colour_distiller(palette = "RdYlBu", name = "capped CV(CD3e)") +
  theme_classic() +
  labs(x = "Mean CD3e", y = "Mean CD20", title = "All cells coloured by capped CV(CD3e)")

tmp |>
  # filter(cell_label == "other") |>
  ggplot(aes(x = mean_CD3e, y = mean_CD20, 
             colour = capVal(fano_CD3e))) +
  geom_point(alpha = 1, size = 1.2) +
  scale_colour_distiller(palette = "RdYlBu", name = "capped Fano(CD3e)") +
  theme_classic() +
  labs(x = "Mean CD3e", y = "Mean CD20", title = "All cells coloured by capped Fano(CD3e)")
```

mean vs variance

```{r}
CD3e = assay(damondSPE, "counts")["CD3e", ]
CD3evar = assay(damondSPE, "var")["CD3e", ]
CD20 = assay(damondSPE, "counts")["CD20", ]

df = data.frame(CD3e, var, CD20)

ggplot(df, aes(x = log10(CD3e+1), y = log10(var+1), colour = sqrt(log10(CD20+1)))) +
  geom_point()
```

```{r}
ggplot(df, aes(x = log10(CD3e+1), y = log10(CD20+1), colour = log10(CD3evar+1)/log10(CD3e+1))) +
  geom_point()
```

```{r}
CD3e = assay(damondSPE, "counts")["CD3e", ]
CD3evar = assay(damondSPE, "var")["CD3e", ]
CD20 = assay(damondSPE, "counts")["CD20", ]
INS = assay(damondSPE, "counts")["INS", ]
celltypes = damondSPE$cell_type

df = data.frame(INS, var, CD3e, celltypes)

ggplot(df, aes(x = log10(INS+1), y = log10(CD3e+1), colour = celltypes)) +
  geom_point()
```

```{r}
rownames(damondSPE)
```


```{r}
table(damondSPE$cell_type)
```


```{r}
means = assay(damondSPE, "counts")["CD20", ]
var = assay(damondSPE, "var")["CD20", ]

plot(log10(means+1), log10(var+1))
```

model mean-variance relationship, anything outside of the expected mean-variance relationship 

Cells with high mean CD3e (whether variance is low or high) - indicates a true CD3e+ cell.

Cells with low mean CD3e and high variance of CD3e - point to CD3e spillover.

Cells with low mean CD3e and low variance of CD3e - point to truly CD3e- cells. 


```{r}
ggplot(tmp, aes(x = mean_CD3e, y = mean_CD20, colour = var_CD20_cap)) +
  geom_point(alpha = 0.6, size = 1.2) +
  scale_colour_distiller(palette = "RdYlBu", name = "Var(CD20)") +
  theme_classic() +
  labs(x = "Mean CD3e", y = "Mean CD20", title = "All cells coloured by variance of CD20 (capped)")

tmp |>
  # filter(cell_label == "other") |>
  ggplot(aes(x = mean_CD3e, y = mean_CD20, 
             colour = capVal(cv_CD20))) +
  geom_point(alpha = 1, size = 1.2) +
  scale_colour_distiller(palette = "RdYlBu", name = " capped CV(CD20)") +
  theme_classic() +
  labs(x = "Mean CD3e", y = "Mean CD20", title = "All cells coloured by capped CV(CD20)")

tmp |>
  # filter(cell_label == "other") |>
  ggplot(aes(x = mean_CD3e, y = mean_CD20, 
             colour = capVal(fano_CD20))) +
  geom_point(alpha = 1, size = 1.2) +
  scale_colour_distiller(palette = "RdYlBu", name = "capped Fano(CD20)") +
  theme_classic() +
  labs(x = "Mean CD3e", y = "Mean CD20", title = "All cells coloured by capped Fano(CD20)")
```

Cells with high mean CD20 (whether variance is low or high) - indicates a true CD20+ cell.

Cells with low mean CD20 and high variance of CD20 - point to CD20 spillover.

Cells with low mean CD20 and low variance of CD20 - point to truly CD20- cells. 

```{r}
tmp = tmp |> mutate(cell_label = case_when(cell_type == "B" ~ "B",
                                     cell_type %in% c("Tc", "Th") ~ "T",
                                     TRUE ~ "other"))

ggplot(tmp, aes(x = mean_CD3e, y = mean_CD20, colour = cell_label)) +
  geom_point(alpha = 0.6, size = 1.2) +
  theme_classic() +
  labs(x = "Mean CD3e", y = "Mean CD20", colour = "Cell type", title = "Cell type plot")
```

### Correlation between CD3e and CD20

```{r}
df = corMat |> as.data.frame() |> rownames_to_column(var = "cellID")
splits = strsplit(df$cellID, "_")
df$image_name  = sapply(splits, `[`, 1)
df$cell_number = sapply(splits, `[`, 2)

corTmp = merge(tmp, df, by = c("image_name", "cell_number"))
```


```{r}
ggplot(corTmp |> filter(!is.na(corTmp$CD20_vs_CD3e)), 
       aes(x = mean_CD3e, y = mean_CD20, colour = CD20_vs_CD3e)) +
  geom_point() +
  theme_classic() +
  scale_colour_distiller(palette = "RdYlBu", name = "Cor(CD20, CD3e)") +
  labs(x = "Mean CD3e", y = "Mean CD20", title = "All cells coloured by Cor(CD20, CD3e)")
```


#### B cells

```{r}
ggplot(corTmp |> filter(!is.na(corTmp$CD20_vs_CD3e) & cell_type == "B"), 
       aes(x = mean_CD3e, y = mean_CD20, colour = CD20_vs_CD3e)) +
  geom_point() +
  theme_classic() +
  scale_colour_distiller(palette = "RdYlBu", name = "Cor(CD20, CD3e)") +
  labs(x = "Mean CD3e", y = "Mean CD20", title = "B cells coloured by Cor(CD20, CD3e)")
```

#### T cells

```{r}
ggplot(corTmp |> filter(!is.na(corTmp$CD20_vs_CD3e) & cell_type %in% c("Th", "Tc")), 
       aes(x = mean_CD3e, y = mean_CD20, colour = CD20_vs_CD3e)) +
  geom_point() +
  theme_classic() +
  scale_colour_distiller(palette = "RdYlBu", name = "Cor(CD20, CD3e)") +
  labs(x = "Mean CD3e", y = "Mean CD20", title = "T cells coloured by Cor(CD20, CD3e)")
```

```{r}
# filter out cells with high correlation
highCorr = corTmp |> filter(CD20_vs_CD3e > 0.7) |> dplyr::select(c("image_name", "cell_number", "cell_id"))
highCorrSPE = damondSPE |> filter(cell_id %in% unique(highCorr$cell_id))
highCorrSPE
```


```{r}
table(highCorrSPE$cell_type)
```



