---
title: "Conditional Probabilities (Heart)"
format: html
editor: visual
---

## Libraries and set up

```{r message = FALSE, warning = FALSE}
.libPaths("/enna/nobackup/biostat/Rpackages/v4/")
suppressPackageStartupMessages({
  library(spicyR)
  library(SpatialDatasets)
  library(tidyverse)
  library(tidySingleCellExperiment)
  library(scClassify)
  library(caret)
})

devtools::load_all("../scClassifyWeights/")
```

```{r message = FALSE, warning = FALSE}
spe = SpatialDatasets::spe_Amancherla_2025()
```

- what is the change in p-value for that one cell type pair?
- for low accuracy cell types, what is happening to L function? 
- accuracy vs L function 
- check if lambda (expected intensity) has changed
- check Kontextual code - statial package 
- write out the K function - and identify where the weights are being included
- 

## EDA

```{r}
spe$patient_id |> unique()
```

```{r}
spe$Sample |> unique()
```

```{r}
spe$patient_biopsy |> unique()
```


```{r}
# cellular/antibody rejection scores (patient level)
spe$patient_cellular_grading |> unique()
spe$patient_antibody_grading |> unique()
```

```{r}
colData(spe) |> as.data.frame() |> distinct(patient_id, patient_rejection_type)
```

```{r}
colData(spe) |> as.data.frame() |> distinct(patient_id, sample_id, patient_rejection_type, biopsy_rejection_type) |>
  arrange(patient_id)
```

```{r}
spe$biopsy_timing |> unique()
```

```{r}
spe$cav_grade |> unique()
```

```{r}
assayNames(spe)
```

```{r}
hist(assay(spe[1, ], "counts"))
```

```{r}
hist(assay(spe[1, ], "logcounts"))
```

```{r}
hist(assay(spe[1, ], "scaledata"))
```

```{r}
colSums(assay(spe, "counts")) |> hist()
```

```{r}
spe$ct_second_pass = as.character(spe$ct_second_pass)
spe$ct_second_pass |> unique()
```

```{r}
# whether rejection responded to immunotherapy
spe$resolution_broad |> unique()
```

```{r}
spe$lineage |> unique()
```

```{r}
colData(spe) |> as.data.frame() |> distinct(patient_id, patient_rejection_type) |> dplyr::count(patient_rejection_type)
```


## No weights spicyR

```{r}
set.seed(51773)

patient_idx = spe |>
  filter(biopsy_timing == "post" & patient_rejection_type == "cellular_rejection") |>
  group_by(resolution_broad) |>
  filter(patient_id %in% sample(unique(patient_id), min(10, n_distinct(patient_id)))) |>
  ungroup() |> pull(patient_id) |> unique()

speSub = spe[, colData(spe)$patient_id %in% patient_idx, ]
```

```{r}
colData(speSub) |> as.data.frame() |> distinct(patient_id, resolution_broad) |> dplyr::count(resolution_broad)
```

```{r}
colData(speSub) |> as.data.frame() |> count(Sample) |> arrange(n)
```

```{r}
speSub$patient_id = droplevels(speSub$patient_id)
speSub$Sample = droplevels(speSub$Sample)
```

```{r}
spicyNoWeights = spicy(speSub,
                       condition = "resolution_broad",
                       subject = "patient_id",
                       imageID = "Sample",
                       cellType = "ct_second_pass",
                       spatialCoords = c("adj_x_centroid", "adj_y_centroid"),
                       window = "convex",
                       cores = 20)
```

```{r}
topPairs(spicyNoWeights)
```

## Generate weights with scClassify

```{r eval = F}
# extract log-normalised expression matrix and existing cell type labels
exprMat = assay(speSub, "logcounts") |> as ("dgCMatrix")
labels  = speSub$ct_second_pass

# train
model = train_scClassify(exprsMat_train = exprMat,
                         cellTypes_train = labels,
                         tree = "HOPACH",
                         selectFeatures = "limma",
                         parallel = TRUE,
                         verbose = TRUE,
                         BPPARAM = BiocParallel::MulticoreParam(workers = 20))

saveRDS(model, "data/scClassifyModel.rds")
```


```{r eval = F}
# extract the wkNN scores (both weights assigned to each cell and the correlation)
# we get 10 weights/correlation values for each cell (k = 10)

devtools::load_all("../scClassifyWeights/")

cellIdx = split(seq_len(ncol(speSub)), cut(seq_len(ncol(speSub)), breaks = 5))
names(cellIdx) = paste0("fold", seq_along(cellIdx))

levelModel = model$hierarchyKNNRes[["limma"]]$model
levelHVG   = model$hierarchyKNNRes[["limma"]]$hvg

chunkSize = 1000
weightResults = list()

for (foldName in names(cellIdx)) {
  
  cat("Processing", foldName, "\n")
  exprMatTest = assay(speSub[, cellIdx[[foldName]]], "logcounts") |> as("dgCMatrix")
  weightResults[[foldName]] = list()
  
  for (i in seq_along(levelModel)) {
    if (is.null(levelModel[[i]])) next
    
    for (j in seq_along(levelModel[[i]])) {
      
      # skip noModel branches
      if ("character" %in% is(levelModel[[i]][[j]])) next
      
      cat("  level", i, "branch", j, "\n")
      
      # find common genes
      commonHVG  = intersect(rownames(exprMatTest), levelHVG[[i]][[j]])
      exprMatSub = exprMatTest[commonHVG, , drop = FALSE]
      exprMatSub = exprMatSub[Matrix::rowSums(exprMatSub) != 0, , drop = FALSE]
      
      if (nrow(exprMatSub) < 5) next
      
      commonGenes = intersect(rownames(exprMatSub), colnames(levelModel[[i]][[j]]$train))
      exprMatSub  = exprMatSub[commonGenes, , drop = FALSE]
      trainMat    = Matrix::t(levelModel[[i]][[j]]$train[, commonGenes])
      
      # chunk test cells to avoid memory issues
      nCells = ncol(exprMatSub)
      chunks = split(seq_len(nCells), ceiling(seq_len(nCells) / chunkSize))
      
      wknnResList = lapply(chunks, function(idx) {
        
        corMat = calculateSimilarity(
          exprsMat_train = trainMat,
          exprsMat_test  = exprMatSub[, idx, drop = FALSE],
          similarity = "pearson")
        
        WKNNcor(corMat = corMat,
          subLevelModel = levelModel[[i]][[j]],
          cutoff_method = "dynamic",
          k = 10,
          prob_threshold = 0.7,
          cor_threshold_static = 0.5,
          cor_threshold_high = 0.7,
          topLevel = FALSE,
          verbose = FALSE)})
      
      # combine chunks
      weightResults[[foldName]][[paste0("level", i, "_branch", j)]] = list(
        predRes = do.call(c, lapply(wknnResList, `[[`, "predRes")),
        topKNN = do.call(cbind, lapply(wknnResList, `[[`, "topKNN")),
        topKNN_cor = do.call(cbind, lapply(wknnResList, `[[`, "topKNN_cor")),
        topKNN_weight = do.call(cbind, lapply(wknnResList, `[[`, "topKNN_weight")),
        level = i,
        branch = j
      )
    }
  }
}

names(weightResults) = c(paste0("fold", 1:5))
```


```{r}
# obtain vote margin
# vote margin = group test cell neighbours by class, sum the weights within each class
# divide weight of each class by total weight to get vote margin
# high confidence cell would have scores close to 1/0 (all neighbours of the same class)
# low confidence cells would have scores that are all around the middle (0.4/0.5) (neighbours of mixed class)

getVoteMargin = function(topKNN_weight, topKNN) {
  # for each cell (column), compute the vote margin
  vapply(seq_len(ncol(topKNN_weight)), function(i) {
    
    weights = topKNN_weight[, i]
    labels  = topKNN[, i]
    
    # aggregate weights by cell type label
    vote = stats::aggregate(weights, by = list(label = labels), sum)
    
    # remove unassigned votes (-1)
    vote = vote[vote$label != "-1", ]
    
    if (nrow(vote) == 0) return(0)  # all neighbours were unassigned
    
    # winning class proportion of total weight
    max(vote$x / sum(vote$x))
    
  }, numeric(1L))
}

# extract vote margins 
confidenceDf = list()

for (foldName in names(weightResults)) {
  for (branchName in names(weightResults[[foldName]])) {
    
    res = weightResults[[foldName]][[branchName]]
    
    voteMargin = getVoteMargin(res$topKNN_weight, res$topKNN)
    
    confidenceDf[[paste0(foldName, "_", branchName)]] = data.frame(
      cellID = names(res$predRes),
      fold = foldName,
      branch = branchName,
      level  = res$level,
      predRes = res$predRes,
      voteMargin = voteMargin
    )
  }
}

confidenceDf = do.call(rbind, confidenceDf)
rownames(confidenceDf) = NULL

# fix the cellID column - strip the leading "1." or any "digits." prefix
confidenceDf$cellID = gsub("^[0-9]+\\.", "", confidenceDf$cellID)

str(confidenceDf)
```

```{r}
head(confidenceDf)
```

```{r}
# how many unique cells (this matches the original SPE object, phew)
length(unique(confidenceDf$cellID))
```

```{r}
# obtain final vote margins (this is the leaf score)
finalConfidence = confidenceDf |>
  dplyr::group_by(cellID) |>
  dplyr::summarise(finalLabel = predRes[which.max(level)],
    nLevels = dplyr::n(),
    confidenceScore = voteMargin[which.max(level)])
```


```{r eval = F}
# obtain predicted cell type labels
# predict using the trained model
nFolds = 5
cellIdx = split(seq_len(ncol(speSub)), cut(seq_len(ncol(speSub)), breaks = nFolds))
names(cellIdx) = paste0("fold", seq_along(cellIdx))

predResults = lapply(cellIdx, function(idx) {
  exprMatTest = assay(speSub[, idx], "logcounts") |> as("dgCMatrix")
  
  pred = predict_scClassify(exprsMat_test = exprMatTest,
                            trainRes = model,
                            algorithm = "WKNN",
                            similarity = "pearson",
                            cutoff_method = "dynamic",
                            parallel = TRUE,
                            BPPARAM = BiocParallel::MulticoreParam(workers = 5),
                            verbose = TRUE)
  return(pred)
})
names(predResults) = paste0("fold", seq_along(predResults))

# extract predicted labels across all folds
predLabelsDf = do.call(rbind, lapply(names(predResults), function(foldName) {
  data.frame(
    cellID = names(predResults[[foldName]]$pearson_WKNN_limma$predRes),
    cellType = predResults[[foldName]]$pearson_WKNN_limma$predRes,
    fold = foldName)})) 

rownames(predLabelsDf) = NULL

# join to finalConfidence
finalConfidence = finalConfidence |>
  dplyr::left_join(predLabelsDf, by = c("cellID"))
```



```{r}
hist(finalConfidence$confidenceScore)
```

## Check scClassify performance

```{r}
# ORDER HAS CHANGED
finalConfidence = merge(finalConfidence, by.x = "cellID",
                        colData(speSub) |> as.data.frame(), by.y = "X")

finalConfidence = finalConfidence |> dplyr::rename(predLabels = cellType)

# overall accuracy
mean(finalConfidence$ct_second_pass == finalConfidence$predLabels)
```

```{r}
# breakdown by cell type... yikes
finalConfidence |>
  dplyr::group_by(ct_second_pass) |>
  dplyr::summarise(n = dplyr::n(),
    accuracy = mean(ct_second_pass == predLabels)) |>
  dplyr::arrange(accuracy)
```


## Weights spicyR

```{r}
cData = colData(speSub) |> as.data.frame()
cData = merge(cData,
              finalConfidence |> dplyr::select(c("cellID", "confidenceScore", "predLabels")),
              by.x = "X", by.y = "cellID")

colData(speSub) = cData |> DataFrame()
```

```{r}
devtools::load_all("../spicyR_shreya/")

spicyWeights = spicyR::spicy(speSub,
                       condition = "resolution_broad",
                       subject = "patient_id",
                       imageID = "Sample",
                       cellType = "ct_second_pass",
                       spatialCoords = c("adj_x_centroid", "adj_y_centroid"),
                       window = "convex",
                       cores = 20)
```

```{r}
topPairs(spicyWeights, n = 700)
```



```{r}
topPairs(spicyNoWeights)
```

```{r}
spicyWeights$p.value["Pericytes__Proliferating endothelial", ]
```


```{r fig.width = 12, fig.height = 8}
signifPlot(spicyNoWeights)
```

```{r fig.width = 12, fig.height = 8}
signifPlot(spicyWeights)
```

```{r}
weightsSig = spicyWeights$p.value |> filter(conditionresolved < 0.05) |> rownames()
noWeightsSig = spicyNoWeights$p.value |> filter(conditionresolved < 0.05) |> rownames()
intersect(weightsSig, noWeightsSig)
```


```{r}
# pairs only in weighted
setdiff(weightsSig, noWeightsSig)
```

```{r}
# pairs only in no weights
setdiff(noWeightsSig, weightsSig)
```

```{r}
spicyBoxPlot(spicyNoWeights, from = "Pericytes", to = "Proliferating endothelial")
```

```{r}
spicyBoxPlot(spicyWeights, from = "Pericytes", to = "Proliferating endothelial")
```


